---
title: "Practical Session Instructions"
author: "Stefano De Sabbata"
date: "24 October 2018"
output: pdf_document
---

# Libraries

This practical session will make use of the "Tidyverse" libraries. If not already installed, please install Tidyverse before cloning the repository, using the command below or via the *Tools > Install Packages* menu in RStudio.

```{r, eval=FALSE}
install.packages("tidyverse")
```



# Repository

The term "repository" refers to the whole collection of code, data, and other files that compose a project, stored on a version-control system. This project is available on [my GitHub ReproducibleResearch repository](https://github.com/sdesabbata/ReproducibleResearch). Download the zipped repository from BlackBoard or GitHub, or clone the repository using *git* if you are familiar with the tool.



## R Markdown

The main tool used to create this reproducible lecture and practical on reproducibility is RMarkdown. That is an R library that allows you to create scripts that mix the Markdown mark-up language and R, to create dynamic documents. RMarkdown script can be compiled, at which point, the Markdown notation is interpreted to create the output files, while the R code is executed and the output incorporated in the document.

The core Markdown notation used in this session is presented below and its interpretation when compiled is further below.

```mark
# Header 1
## Header 2
### Header 3
#### Header 4
##### Header 5

**bold**
*italics*

[This is a link to the University of Leicester](http://le.ac.uk)

- Example list
    - Main folder
        - Analysis
        - Data
        - Utils
    - Other bullet poit
- And so on
    - and so forth
```

# Header 1
## Header 2
### Header 3
#### Header 4
##### Header 5

**bold**
*italics*

[This is a link to the University of Leicester](http://le.ac.uk)

- Example list
    - Main folder
        - Analysis
        - Data
        - Utils
    - Other bullet poit
- And so on
    - and so forth

## Build

To build all the scripts in the repository in the correct order, please execute the *Make.R* script that you can find in the main folder.

```{r, eval=FALSE}
source('Make.R')
```