---
title: "Practical Session Instructions"
author: "Stefano De Sabbata"
date: "24 October 2018"
output: pdf_document
---

# Libraries

This practical session will make use of the "Tidyverse" libraries. If not already installed, please install Tidyverse before cloning the repository, using the command below or via the *Tools > Install Packages* menu in RStudio.

```{r, eval=FALSE}
install.packages("tidyverse")
```



# Markdown

The main tool used to create this reproducible lecture and practical on reproducibility is [RMarkdown](https://rmarkdown.rstudio.com/) That is an R library that allows you to create scripts that mix the Markdown mark-up language and R, to create dynamic documents. RMarkdown script can be compiled, at which point, the Markdown notation is interpreted to create the output files, while the R code is executed and the output incorporated in the document.

The core Markdown notation used in this session is presented below and its interpretation when compiled is further below.

```mark
# Header 1
## Header 2
### Header 3
#### Header 4
##### Header 5

**bold**
*italics*

[This is a link to the University of Leicester](http://le.ac.uk)

- Example list
    - Main folder
        - Analysis
        - Data
        - Utils
    - Other bullet poit
- And so on
    - and so forth
```

# Header 1
## Header 2
### Header 3
#### Header 4
##### Header 5

**bold**
*italics*

[This is a link to the University of Leicester](http://le.ac.uk)

- Example list
    - Main folder
        - Analysis
        - Data
        - Utils
    - Other bullet poit
- And so on
    - and so forth



## R Markdown

R code can be embedded in RMarkdown documents as in the example below. That will result in the code chunk be displayed within the document (as *echo=TRUE* is specified), followed by the output from the execution of the same code.

````
`r ''````{r, echo=TRUE}
for (i in 1:4) {
    if (i %% 2 == 0){
        cat("even \n")
    } else {
        cat("odd \n")
    }
}
`r ''````
````

```{r, echo=TRUE}
for (i in 1:4) {
    if (i %% 2 == 0){
        cat("even \n")
    } else {
        cat("odd \n")
    }
}
```



# Part 1

To create an RMarkdown document in RStudio, select *File > New File > R Markdown ...*. On the RMarkdown document creation menu, specify a title and your name as author, and select *PDF* as default output format. The new document should contain only the core document information, as in the example below.

```
---
title: "Practical Session Instructions"
author: "Stefano De Sabbata"
date: "24 October 2018"
output: pdf_document
---
```

Copy the following text below the document information.

````
# A nice heading

This is my first [RMarkdown](https://rmarkdown.rstudio.com/) document.

The code below will print:

- if the number is even
    - even
- else
    - odd

`r ''````{r, echo=TRUE}
for (i in 1:4) {
    if (i %% 2 == 0){
        cat("even \n")
    } else {
        cat("odd \n")
    }
}
`r ''````
````




# Part 2

## Repository

The term "repository" refers to the whole collection of code, data, and other files that compose a project, stored on a version-control system. This project is available on [my GitHub ReproducibleResearch repository](https://github.com/sdesabbata/ReproducibleResearch). Download the zipped repository from BlackBoard or GitHub, or clone the repository using *git* if you are familiar with the tool.



## Build

To build all the scripts in the repository in the correct order, please execute the *Make.R* script that you can find in the main folder.

```{r, eval=FALSE}
source('Make.R')
```