---
title: "Practical Session Instructions"
author: "Stefano De Sabbata"
date: "24 October 2018"
output: pdf_document
---

# Libraries

As mentioned in earlier lectures, libraries are collections of functions. Libraries can be installed in R using the function *install.packages* or in RStudio via the *Tools > Install Packages* menu. This practical session will make use of the "Tidyverse" libraries, that includes a number of different libraries, including *knitr*. The *Make.R* script includes the instruction to install the Tidyverse if not currently installed. You can check whether the Tidyverse is installed by clicking on the *Packages* tab in the bottom-right pane in RStudio, and search for "tidyverse". If not installed, install it using the command below or via the *Tools > Install Packages* menu in RStudio.

```{r, eval=FALSE}
install.packages("tidyverse")
```

You can load a library using the function *library*, as shown below. Once a library is installed on a computer you don't need to install it again, but every script needs to load all the library that it uses. Once a library is loaded all its functions can be used.



# Markdown

The main tool used to create this reproducible lecture and practical on reproducibility is [RMarkdown](https://rmarkdown.rstudio.com/) That is an R library that allows you to create scripts that mix the Markdown mark-up language and R, to create dynamic documents. RMarkdown script can be compiled, at which point, the Markdown notation is interpreted to create the output files, while the R code is executed and the output incorporated in the document.

The core Markdown notation used in this session is presented below and its interpretation when compiled is further below.

```mark
# Header 1
## Header 2
### Header 3
#### Header 4
##### Header 5

**bold**
*italics*

[This is a link to the University of Leicester](http://le.ac.uk)

- Example list
    - Main folder
        - Analysis
        - Data
        - Utils
    - Other bullet poit
- And so on
    - and so forth
```

# Header 1
## Header 2
### Header 3
#### Header 4
##### Header 5

**bold**
*italics*

[This is a link to the University of Leicester](http://le.ac.uk)

- Example list
    - Main folder
        - Analysis
        - Data
        - Utils
    - Other bullet point
- And so on
    - and so forth



## R Markdown

R code can be embedded in RMarkdown documents as in the example below. That will result in the code chunk be displayed within the document (as *echo=TRUE* is specified), followed by the output from the execution of the same code.

````
`r ''````{r, echo=TRUE}
for (i in 1:4) {
    if (i %% 2 == 0){
        cat("even \n")
    } else {
        cat("odd \n")
    }
}
`r ''````
````

```{r, echo=TRUE}
for (i in 1:4) {
    if (i %% 2 == 0){
        cat("even \n")
    } else {
        cat("odd \n")
    }
}
```



## Other necessary software

In order to compile the RMarkdown scripts to *.pdf* documents, RMarkdown requires a LaTeX compiler. 

If you are working on the [University of Leicester](https://le.ac.uk/) Windows 10 computers, please check that [MiKTeX](https://miktex.org/) is installed in the *Software Centre*. If not close RStudio if open and install it (this might take a few minutes). Once the installation is completed, re-open RStudio. 

MiKTeX should work fine for any other Windows system. Make sure to select "Always install missing packages on-the-fly" in the general settings of the MiKTex console, so that the appropriate packages are installed when the *.pdf* document is compiled. On Linux, please install a TeX distribution, e.g. `sudo apt-get install texlive-full` on Ubuntu. On Mac OS, please install a TeX distribution such as [MacTeX](http://www.tug.org/mactex/).

In RStudio, select *Tools > Global Options...*. In the *Swave* section, make sure that the following options are selected.

- Weave Rnw files using: *knitr*
- Typeset LaTeX into PDF using: *XeLaTeX*



# Part 1

To create an RMarkdown document in RStudio, select *File > New File > R Markdown ...*. On the RMarkdown document creation menu, specify a title and your name as author, and select *PDF* as default output format. The new document should contain only the core document information, as in the example below.

```
---
title: "Practical Session Instructions"
author: "Stefano De Sabbata"
date: "24 October 2018"
output: pdf_document
---
```

Copy the following text below the document information and the click on the *Knit* button on the bar above the editor panel (top-left area) in RStudio, on the left side. Check the resulting *pdf* document. Try add some of your own code (e.g., from previous practical sessions) and Markdown text, and compile the document again.

````
# A nice heading

This is my first [RMarkdown](https://rmarkdown.rstudio.com/) document.

The code below will print:

- "even"
    - if the number is even
- "odd"
    - otherwise

`r ''````{r, echo=TRUE}
for (i in 1:4) {
    if (i %% 2 == 0){
        cat("even \n")
    } else {
        cat("odd \n")
    }
}
`r ''````
````



# Part 2



## Repository

The term "repository" refers to the whole collection of code, data, and other files that compose a project, stored on a version-control system. This project is available on [my GitHub ReproducibleResearch repository](https://github.com/sdesabbata/ReproducibleResearch). Download the zipped repository from BlackBoard or GitHub, or clone the repository using *git* if you are familiar with the tool.

Once downloaded and un-zipped in a folder of your choice, observe the structure structure of the folder and sub-folders. The *Data* folders contains two scripts that will download some data in the same folder. A back-up version of those data can be found in the *Backup* folder. The *Analysis* folder contains a simple analysis of those data. The *Materials* section contains the lecture and practical session materials, and some additional folders containing the [IOSlides templates](https://bookdown.org/yihui/rmarkdown/ioslides-presentation.html) and the images. The *Utils* folder contains a script to update the git branch connected to the [GitHub Pages](https://pages.github.com/) for [this project](https://sdesabbata.github.io/ReproducibleResearch/).



## Load

In RStudio, click on *File > Open Project...* and navigate to the folder where the files have been un-zipped. Select the file *ReproducibleResearch.Rproj* (the extension *.Rproj* might be hidden on Windows, the file type is *R Project*) and click *Open*. RStudio will load the whole project and set the directory where *ReproducibleResearch.Rproj* is located as working directory.



## Build

The *Make.R* script in the main folder can be used to "build" all the scripts in the repository in the correct order. The *Make_Clean.R* script can be used to delete all the files created by the *Make.R*.

Build the project by running the *Make.R* script. That can be done by loading the file in RStudio and clicking on the *Source* button on bar above the editor panel (top-left area) in RStudio, right side.

Alternatively, you can run the *Make.R* file using the command below from the RStudio console.

```{r, eval=FALSE}
source('Make.R')
```



# Part 3

Once the project has been built, the *Reproducible_analysis_in_R.pdf* file will be available in the *Analysis* folder. Open the document, read and complete the exercise.



## Close

Once completed the practical, click on the project name *ReproducibleResearch* in the top-left corner in RStudio and then *Close project* to close the project. RStudio might show a prompt asking whether to save the *.RData* file before closing, in which case you can select to close without saving.